### 2023-06-04 manuscript/chapters/2_neural_fields.qmd
this takes so much time nyaaa

@@

no external current,

general low-rank neural fields, F and G notation

$$
J_{ij} = \sum_{\mu=1}^p F_{\mu,i} G_{\mu,j}
$$

show how the embedding becomes smooth

@@

- $h_i = \xi_{\mu,i} \quad \forall \mu = 1,\cdots,p$ (stable)
- $h_i=0$ (unstable)

@@

## Building up to a neural field in $\mathbb R^2$

@@

$$
\begin{aligned}
\partial_t h(t, \vec z) &= -h(t, \vec z) + \int_{\mathbb{R}^p} w(\vec z, \vec y) \phi(h(t, \vec y)) \rho(\mathrm d \vec y),
\quad
w(\vec z, \vec y) = \sum_{\mu=1}^p \tilde \phi (y_\mu) z_\mu \\
&= -h(t, \vec z) + \sum_{\mu=1}^p z_\mu m_\mu(t)
\end{aligned}
$$

study of fixed points (and convergence of the fixed points)

@@

@@

@@

introduce overlap, and in our case the low-rank model the kappa

overlaps as an approximation of the “dynamics” (latent space), in the case phi=linear

The dynamics of $h(t, \vec z)$ are in a subsystem of dimension $p$ spanned by the ONB of functions $\{e_\mu(\vec z) = z_\mu | \mu=1,\cdots,p\}$, with the scalar product $\langle f, g \rangle = \int_{\mathbb R^p} f(\vec y) g(\vec y) \rho(\mathrm d \vec y)$.

We decompose $h(t, \vec z) = h^\perp(t, \vec z) + \sum_{\mu=1}^p \kappa_\mu(t) z_\mu$, and write the system of equations for $\mu=1,\cdots,p$.

$$
\dot \kappa_\mu(t) = -\kappa_\mu(t) + \int_{\mathbb{R}^p} \tilde\phi(y_\mu) \phi(h(t, \vec y)) \rho(\mathrm d \vec y) = -\kappa_\mu(t) + m_\mu(t)
$$

with initial conditions

$$
\kappa_\mu(0) = \int_{\mathbb{R}^p} y_\mu h(0, \vec y) \rho(\mathrm d \vec y)
$$

and the orthogonal component evolves according to

$$
h^\perp(t, \vec z) = h^\perp(0, \vec z) \mathrm e^{-t},
\quad
h^\perp(0, \vec z) = h(0, \vec z) - \sum_{\mu=1}^p \kappa_\mu(0) z_\mu
$$

@@

@@

show what the analytical kappa is

introduce both, derive kappa again

@@

recurrent current picture

>>>>

::: {.content-hidden}
{{< include ../macros.tex >}}
:::

@@

### Case study : low-rank networks of neurons

start with the study of network of neurons, no external current

[TODO MATH : show self-connections $\sim \frac 1N$, so we ignore them]

general setting of low-rank networks

$$
\begin{aligned}
\dot h_i(t) = -h_i(t) + \frac 1N \sum_{\mu=1}^p \sum_{j=1}^N F_{\mu i} G_{\mu j} \phi(h_j(t)) \\
F_{\mu i} = f_\mu(z_{1 i}, \cdots, z_{p i}),\, G_{\mu i} = g_\mu(z_{1 i}, \cdots, z_{p i}),\, \vec{z_i} \stackrel{\text{iid}}{\sim} \rho
\end{aligned}
$$

interpretation : each neuron has a position $\vec{z_i} \in \Rp$. $J_{ij} = \frac 1N \sum_{\mu=1}^p = F_{\mu i} G_{\mu j}$ is the weight of recurrent currents.

our setting is :

$$
\begin{aligned}
f_\mu(z_1, \cdots, z_p) = z_\mu \\
g_\mu(z_1, \cdots, z_p) = \tilde \phi(z_\mu) = \frac{\phi(z_\mu) - \avg{\phi(z_\mu)}}{\mathrm{Var}[\phi(z_\mu)]} \\
\phi(z) = \frac{1}{1+\mathrm{e}^{-z}} \\
\vec{z} = (z_1, \cdots, z_p) \sim \rho(z_1, \cdots, z_p) = \prod_{\mu=1}^p \mathcal{N}(z_\mu),\, \mathcal{N}(z) = \frac{1}{\sqrt{2 \pi}} \mathrm{e}^{-\frac 12 z^2} \\
\end{aligned}
$$

@@

* $h_i = z_{\mu,i} \quad \forall \mu = 1,\cdots,p$ (stable)
* $h_i=0$ (unstable)

@@

talk about since $\lambda_\nu = \avg{g^\prime_\nu(z_\nu) f_\nu(z_\nu)}-1$, the sharper the gain, the more stable to fixed points (since the bump of $\delta \phi$ becomes more concentrated)

[TODO FIG : draw a flow field of in the latent space]
[TODO FIG : numerical eigenvalues plot]

### Emergence of structure in the networks of neurons

show simulations, first in $p=1$, then $p=2$.

![simulation in 1D, p=1](figures/embedding_d=1_h0=0_phi=sigmoid.mp4)

![simulation in 2D, p=2](figures/embedding_d=2_h0=0.mp4)

[TODO FIG : scatterplot of the network, with animation]

show it converges to the fixed points $z_1$ and $z_2$

the structure of the scatterplot emerges naturally as a smooth manifold in $\R^2$

### Direct simulation of the low-rank neural field

@@

$$
\begin{aligned}
\partial_t h(t, \vec z) &= -h(t, \vec z) + \int_{\mathbb{R}^p} w(\vec z, \vec y) \phi(h(t, \vec y)) \rho(\mathrm d \vec y),
\quad
w(\vec z, \vec y) = \sum_{\mu=1}^p \tilde \phi (y_\mu) z_\mu
\end{aligned}
$$

repeat the fixed points are now the generalization, functions $h=z_1$ and $h=z_2$

@@

[TODO : show the connectivity with grid sampling. mention that the problems with the compact support get adressed later in the next chapter, when we do $\R^2 \mapsto [0,1]^2$]

@@

note that :

$$
\partial_t h(t, \vec z) = -h(t, \vec z) + \sum_{\mu=1}^p z_\mu m_\mu(t)
$$

[TODO : i'm not even sure that introducing the overlaps in necessary. we might just include them for completeness, and because they are relevant in papers on low-rank stuff]

@@

introduce kappa, and in our case the low-rank model the kappa

we define

$$
\kappa_\mu(t) = \avg{f_\mu, h(t, \cdot)} = \int_{\Rp} y_\mu h(0, \vec y) \rho(\mathrm d \vec y)
$$

with these definition, we have in our case
$$
\begin{aligned}
\dot \kappa_\mu(t) &= -\kappa_\mu(t) + \int_{\Rp} \tilde\phi(y_\mu) \phi(h(t, \vec y)) \rho(\mathrm d \vec y) = -\kappa_\mu(t) + m_\mu(t) \\
\kappa_\mu(0) &= \int_{\mathbb{R}^p} y_\mu h(0, \vec y) \rho(\mathrm d \vec y)
\end{aligned}
$$

overlaps as an approximation of the “dynamics” (latent space), in the case $\phi=\mathrm{linear}$. since in our case $\phi$ is a sigmoid, it can be considered linear in the range $h \sim \pm 3$. we therefore use the computed overlaps as to show the latent space

[TODO MATH : actually we could also juste compute the kappa as a function of time, and study that]

@@

we introduce $\delta$ the delay, and rolling $\mu+1$, with convention $p+1=1$

$$
\partial_t h(t, \vec z) = -h(t, \vec z) + \int_{\mathbb{R}^p} \sum_{\mu=1}^p \tilde \phi (y_\mu) z_{\mu+1} \phi(h(t - \delta, \vec y)) \rho(\mathrm d \vec y)
$$

note this is now a DDE instead of a PDE

@@

[TODO : is this useful ? I derived this because it shows we need both delay and rolling, but might be too much]

we can show that in this case

$$
\dot \kappa_\mu(t) = -\kappa_\mu(t) + m_{\mu+1}(t - \delta)
$$

then in the case that $\phi=linear \implies \kappa_\mu = m_\mu$, and initial conditions $m_\mu(t<0) = m_\mu(0)$ (constant inititial condition), $m_1(0) = 1$ and $m_2(0) = 0$, we can solve the first "cycle" ($0 < t < \delta$) analytically :

$$
\begin{aligned}
\kappa_1(t) &= \mathrm{e}^{-t} \\
\kappa_2(t) &= 1 - \mathrm{e}^{-t} \\
\end{aligned}
$$

then one can repeat this for the next cycles. basically each cycle is applying an exponential convolution filter

[TODO MATH : the derivation of this, we have an exponential convolution filter]

![](figures/kappa_cycling_p=2.tmp.png)

@@

[TODO : might be good to show a simulation where we explicitly simulate $\kappa$ ? but the goal here is just to "prove" the cycling behavior]

![](figures/embedding_d=2_h0=0_cycling_delta=6.mp4)

![Intuition for the cycling behavior by plotting the recurrent currents](figures/cycling_recurrent.png)

## From $\R^2$ to $[0,1]^2$

Defining the change of variables $u_\mu=\mathrm{CDF}(y_\mu), v_\mu=\mathrm{CDF}(z_\mu)$, the neural field becomes

$$
\begin{aligned}
& \partial_t h_U(t, \vec v) = -h_U(t, \vec v) + \int_{[0,1]^p} w_U(\vec v, \vec u) \phi(h_U(t, \vec u)) \mathrm d \vec u \\
& \text{where } w_U(\vec v, \vec u) = w(\mathrm{CDF}^{-1}(\vec v), \mathrm{CDF}^{-1}(\vec u)), \, h_U(t, \vec v) = h(t, \mathrm{CDF}^{-1}(\vec v))
\end{aligned}
$$

![normal distribution CDF](figures/1920px-Normal_Distribution_CDF.svg.png)

We now have a neural field equation on a unit (hyper)cube, and uniform sampling.

the problem of sampling all of $\R^2$ is now "hidden away" in the CDF, since it goes to $\pm \infty$ at 0 and 1.

note : here we use the fact that the components are independant. therefore we can define the total as the product, i.e.

$$
\mathrm{CDF}(z_1, z_2) = \int_{-\infty}^{z_1} \rho_1(d y_1) \int_{-\infty}^{z_2} \rho_2(\d y_2) = v_1 v_2
$$

we would not be able to do this is the coordinates were coupled, then we could define still define the CDF, but we cannot factorize it into 2 different components

![](figures/sampling_R2_01.png)

[TODO : does this deserve its own section ? maybe mention it later, we can give a picture of the "isolines" of the 2D gaussian CDF]

## Numerical simulation of neural fields

### Direct sampling

we simulate the network of neurons

### Bin averaging samples

![](figures/sampling_R2_nogrid_kernel.png)

### Grid over $[0,1]^2$

We define $\vec z(\alpha) = S(\alpha)$ the 2D point corresponding to the mapping $\alpha$. (details : we need a bounding box to map back). Defining a matrix $Z_{\mu,\alpha} = \vec z(\alpha)_\mu$, we can write down a numerical PDF $\tilde \rho(Z_{1,\alpha},\cdots,Z_{p,\alpha}) = \tilde \rho(Z_{:,\alpha}) = \frac{\rho(Z_{:,\alpha})}{\sum_\beta \rho(Z_{:,\beta})}$ and the following patterns to simulate the embedded $[0,1]$ neural field as a low-rank RNN.

$$
\tilde F_{\mu,\alpha} = Z_{\mu,\alpha}, \quad \tilde G_{\mu,\alpha} = \tilde\phi(Z_{\mu,\alpha}), \quad \tilde J_{\alpha,\beta}=\tilde \rho(Z_{:,\beta}) \sum_{\mu=1}^p \tilde F_{\mu,\alpha} \tilde G_{\mu,\beta}
$$

Doing this is equivalent to the formulation of the binned connectivity matrix.

![](figures/sampling_R2_01_kernel.png)

### 2023-06-04 manuscript/chapters/3_mappings.qmd
this takes so much time nyaaa

@@

@@

Define a measurable mapping $S : \mathbb{R}^p \rightarrow [0, 1]$.

Let $\mu : [0,1] \rightarrow [0, \infty]$ be a measure on $([0,1], \mathcal B([0,1]))$.

Then we can write

$$
\partial_t h(t, \vec z) = -h(t, \vec z) + \int_{[0,1]} [w(\vec z, \cdot) \phi(h(t, \cdot)) \rho(\cdot)] \circ S^{-1} \; \mathrm d \mu
$$

## From $\R^2$ to $[0,1]^2$

Defining the change of variables $u_\mu=\mathrm{CDF}(y_\mu), v_\mu=\mathrm{CDF}(z_\mu)$, the neural field becomes

$$
\partial_t h_U(t, \vec v) = -h_U(t, \vec v) + \int_{[0,1]^p} w_U(\vec v, \vec u) \phi(h_U(t, \vec u)) \mathrm d \vec u, \quad
w_U(\vec v, \vec u) = w(\mathrm{CDF}^{-1}(\vec v), \mathrm{CDF}^{-1}(\vec u)), \quad h_U(t, \vec v) = h(t, \mathrm{CDF}^{-1}(\vec v))
$$

We now have a neural field equation on a unit (hyper)cube, and uniform sampling.

@@

linear projections don’t work ofc

@@

@@

@@

argue over binary expansion

@@

how to we actually simulate the mappings in one-dimension

how to we actually simulate the mappings in one-dimension

@@

@@

## Reinterpretation of simulation strategies

### Bin averaging samples

give an intuition, show the figure and some simulations to show that it works

### Direct remapping of the grid

We define $\vec z(\alpha) = S(\alpha)$ the 2D point corresponding to the mapping $\alpha$. (details : we need a bounding box to map back). Defining a matrix $Z_{\mu,\alpha} = \vec z(\alpha)_\mu$, we can write down a numerical PDF $\tilde \rho(Z_{1,\alpha},\cdots,Z_{p,\alpha}) = \tilde \rho(Z_{:,\alpha}) = \frac{\rho(Z_{:,\alpha})}{\sum_\beta \rho(Z_{:,\beta})}$ and the following patterns to simulate the embedded $[0,1]$ neural field as a low-rank RNN.

$$
\tilde F_{\mu,\alpha} = Z_{\mu,\alpha}, \quad \tilde G_{\mu,\alpha} = \tilde\phi(Z_{\mu,\alpha}), \quad \tilde J_{\alpha,\beta}=\tilde \rho(Z_{:,\beta}) \sum_{\mu=1}^p \tilde F_{\mu,\alpha} \tilde G_{\mu,\beta}
$$

Doing this is equivalent to the formulation of the binned connectivity matrix.

## Numerical simulations of the neural field in $[0,1]$

@@

averaging of 2ˆn neurons

see sampling.ipynb

>>>>

[TODO MOTIVATION : is this really useful ? the point i'm trying to make here is the embedding problem arises when you have an infinite number]

@@

$$
\begin{aligned}
\rho(\vec z) &= \sum_{a \in P} p_{a} \delta(\vec{z_a} - \vec{z}) \\
\text{normalization : } \int_{\Rp} \rho(\d \vec{z}) &= \sum_{a} p_a = 1
\end{aligned}
$$

we can explicitly formulate the embedding with a new 1D PDF (arrange the populations on the line, their probability is their segment length)

$$
\gamma(\alpha) = \sum_{a \in P} p_a \mathbb{1}(C_{a-1} \leq \alpha < C_{a}),\, C_a=\sum_{b=1}^a p_b
$$

the connectivity matrix in the 1D model becomes

$$
J_{ab} = \sum_{\mu=1}^p f_\mu(\vec{z_a}) g_\mu(\vec{z_b})
$$

[TODO MATH : flesh this out]

case 2 populations, we get the ising model (binary patterns)

the point i try to make is that for discrete distributions (i can enumerate the populations), it is easy to embed in 1D. the problem comes when the distribution is continuous.

this also helps give an intuition that effectively, when we do the binning in $[0,1]^2$, each bin represents a population

@@

Define a measurable and bijective mapping $S : \mathbb{R}^p \rightarrow [0, 1]$.

Let $\lambda : [0,1]^2 \rightarrow [0, \infty]$ be a measure.

We define $\tilde h(t, \alpha) = h_U(t, S^{-1}(\alpha))$, $\tilde w(\alpha, \beta) = w_U(S^{-1}(\alpha), S^{-1}(\beta))$

$$
\partial_t \tilde h(t, \alpha) = -\tilde h(t, \alpha) + \int_{[0,1]} \tilde w(\alpha, \beta) \tilde\phi(h(t, \beta)) \lambda \circ S^{-1}(\d \beta)
$$

![](figures/mappings_h1d_state.png)

![](figures/mapping_recursive_quadrant.mp4)

@@

linear projections don’t work ofc.

Note : why are not using the CDF to embed in 1D ? because it's not bijective !

@@

![](figures/mapping_demo_indices.png)

[TODO : do we mention the random mapping ?]

@@

note that in other definitions (and in code) it is defined by $0.b^y_1 b^x_1 b^y_2 b^x_2 \cdots$. This difference is not fundamentally important though.

@@

argue over binary expansion, and especially how the fact that

* it converges forces the least significant bits of $z_1$ and $z_2$ to go to the lsb of $\alpha$
* it's a bijection means "both bits of x and y" need to be used in the expansion

@@

@@

what we are doing in reality is assigning indices to squares of size $2^{-n} \times 2^{-n}$, and mapping them to indices representing segments of length $4^{-n}$

[TODO FIG : drawing of the square-to-segment mapping, something like

![Alt text](figures/Screenshot_2023-06-03_19-59-49.png)]

@@

## Coarse-graining the neural field in $[0,1]$

@@

In order to simulate the neural field numerically, we discretize by taking a grid of $4^n$ bins in the $[0,1]^2$ space.

$$
\int_{[0,1]^2} w_U(\vec v, \vec u) \phi(h_U(t, \vec u)) \mathrm d \vec u \rightarrow \sum_{j=1}^N \underbrace{w_U(\vec u_i, \vec u_j)}_{J_{ij}} \phi(\underbrace{h_U(t, \vec u_j)}_{h_j(t)})
$$

Each bin therefore corresponds to a population (density) indexed by $i = \{1\cdots4^n\}$. Population $i$ interacts with other populations through the (numerically finite) connectivity $J_{ij}$.

To simulate the neural field in $[0,1]$, we use the mapping $S(\vec u_i) \stackrel{\text{notation}}= S(i) = \alpha_i$ to embed the populations in 1D.

One trivial mapping would be $S(i) = i / 4^n$, which is just ordering the populations in [0,1] in the same order they appear inside the bins. More elaborate mappings are simply *permutations* of the numerically binned populations set $\{1\cdots4^n\}$, and the resulting connectivity matrix is then given by $J_{S(i),S(j)}$ [^1].

However, the connectivity matrix is invariant to permutations of neurons, and without further modification, the dynamics emerging from $J_{S(i),S(j)}$ are the same as the ones emerging from $J_{ij}$.

In the same way that discretizing the smooth 2D field works *because of* the smoothness, we discretize the 1D field emerging from the mapping. If the mapping has "nice properties" (e.g. "locality"), this discretization should work in a similar manner.

For this, we consider the $4^n$ mapped populations $\alpha_i$ to be the "mathematically exact" 1D field, and we "discretize" the 1D field by taking a grid of $2^n$ bins in $[0,1]$. In effect, we average bins of $2^n$ populations, in order to obtain $4^n/2^n = 2^n$ coarse-grained populations in $[0,1]$. We call this "downsampling" [^2].

The main question is now, what conditions must be imposed on the mapping $S : [0,1]^2 \rightarrow [0,1]$ (numerically : a permutation), such that it is *resilient* to this downsampling, in other words that the dynamics obtained from the connectivity in 1D are the same as the dynamics in 2D [^3].

In the following, this assertion is made by comparing the trajectories $(m_0(t),m_1(t))$ in phase space (also called latent space). If the mapping is "resilient", small differences between the trajectories vanish as $n \rightarrow \infty$ [^4].

[^1] : I am glossing over the details of how we bin the grid of $4^n$ elements in $[0,1]^2$ space. Numerically, there is an order in which we enumerate each of the $4^n$ bins (the most obvious is column-wise enumeration, i.e. ``ReshapeMapping.inverse_samples()``). As described, this ordering is abitrary, and one should always consider the enumeration of the bins to be in random order.

[^2] : The reason for which we take bins of size $2^n$, is because $2^n$ is the maximum size a bin can have which spans in only one direction in the 2D embedding. For instance, with the ``ReshapeMapping`` a column contains $2^n$ populations, all with the same $\xi_0$ position : visually the bin is vertical. With the ``DiagonalMapping``, the largest sequence of populations which "go in the same direction" (TODO : clarify this) is of size $2^n$, and spans the diagonal (1,0)->(0,1). Intuitively, this averaging can be seen as "destroying fast and large oscillations".

[^3] : The connectivity $\tilde J_{\alpha\beta}$ in 1D can be obtained by computing the "average low-rank vectors". See ``draft.md`` for more details.

[^4] : Each point in $(m_0,m_1)$ fully encodes the state of a linear low-rank RNN. We can prove (see my own handwritten notes) that $m_\mu(t) \approx \kappa_\mu(t)$ when $\phi(h)$ is close to linear. In our case, $\phi(h)$ is a sigmoid (I found a sigmoid has better convergence properties than a linear function), and can be considered linear as long as $-3 \lesssim h(t) \lesssim 3$.

![](figures/coarsegraining.png)
